---
title: "README"
author: "Aaron"
date: "2023-06-20"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Google Certificate in Data Analytics
## Capstone Project #1

### Introduction
This is the first of several projects for practicing and showcasing what I have learned throughout this course. In this case study, my main focus is to develop a solid R Markdown document for ease of creating a report that is easily followed by colleagues and stakeholders.

#### Notes
These data are provided under [this license](https://ride.divvybikes.com/data-license-agreement). Since the dataset(s) are extremely large, I intend to compile and clean them in BigQuery and/or R. If there ends up being a smaller dataset, I might switch to a spreadsheet, unless I have already laid the necessary infrastructure in R.

## Business Goals (Ask)
Director of Marketing, Lily Moreno has asked for analysis of Cyclistic's internal data, with the overarching goal of converting casual users into annual subscribers. She has asked three specific questions:

* How do annual members and casual riders use Cyclistic bikes differently?

* Why would casual riders buy Cyclistic annual memberships?

* How can Cyclistic use digital media to influence casual riders to become members?

For this report, we will focus on the first question and analyze the differences in use patterns by casual riders and annual members.

## Data Preparation (Prepare/Process)
Ride data from June 2022 to May 2023 were retrieved from Cyclistic's internal database. (Note: the data are actually made available by Motivate International Inc. via the previous link.)

The data files contain tables with about 200,000 to 1,000,000 rows, so Spreadsheets will be too bulky for the initial cleaning and preparation.

#### These are the necessary packages for this analysis.
```{r}
library(tidyverse)
library(readr)
library(lubridate)
library(ggplot2)
library(knitr)
getwd()
setwd("~/Desktop/Coding Projects/Capstone_Cyclistic_Raw_Data/Cyclistic_csv_files")
```

#### Importing the necessary data.
##### Note: changed September from "trippublicdata" to "tripdata" for consistency.
```{r}
setwd("~/Desktop/Coding Projects/Capstone_Cyclistic_Raw_Data/Cyclistic_csv_files")
Jun_2022 <- read_csv("202206-divvy-tripdata.csv")
Jul_2022 <- read_csv("202207-divvy-tripdata.csv")
Aug_2022 <- read_csv("202208-divvy-tripdata.csv")
Sep_2022 <- read_csv("202209-divvy-tripdata.csv")
Oct_2022 <- read_csv("202210-divvy-tripdata.csv")
Nov_2022 <- read_csv("202211-divvy-tripdata.csv")
Dec_2022 <- read_csv("202212-divvy-tripdata.csv")
Jan_2023 <- read_csv("202301-divvy-tripdata.csv")
Feb_2023 <- read_csv("202302-divvy-tripdata.csv")
Mar_2023 <- read_csv("202303-divvy-tripdata.csv")
Apr_2023 <- read_csv("202304-divvy-tripdata.csv")
May_2023 <- read_csv("202305-divvy-tripdata.csv")
```

#### After using colnames() and str() to ensure matching datasets, 
#### the above datasets were merged.
```{r}
all_trips <- rbind(Jun_2022, Jul_2022, Aug_2022, Sep_2022, Oct_2022, 
                       Nov_2022, Dec_2022, Jan_2023, Feb_2023, Mar_2023,
                       Apr_2023, May_2023)
```

#### A quick snapshot for orientation and an idea of scale.
```{r}
str(all_trips)
ncol(all_trips)
nrow(all_trips)
```



#### Confirmation that there are only two labels for members.
```{r}
categories <- unique(all_trips$member_casual)
number_of_categories <- length(categories)
list(categories)
table(all_trips$member_casual)
```
There are no extraneous values to be removed, although there will be further
cleaning of nonsensical values.

#### These next steps split the datetime column into more human time, 
#### day, month, etc. Not entirely sure why format() could not extract the
#### day of the week, but the weekdays() function worked perfectly.
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date),"%m")
all_trips$day <- format(as.Date(all_trips$date),"%d")
all_trips$year <- format(as.Date(all_trips$date),"%Y")
all_trips$day_of_week <- weekdays(as.Date(all_trips$date))
```

#### Confirmation of the correct structure.
```{r}
table(all_trips$month)
table(all_trips$day)
table(all_trips$year)
table(all_trips$day_of_week)
```
12 months, 31 days, 2 years, and 7 days of the week, as expected.

#### Calculating the ride length and ensuring the correct format.
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

#### Need to remove negative trip lengths and quality control checks
#### This is the formula I used, as the method using an operator did not work.
#### After a dplyr update, the ! operator method worked perfectly well.
```{r}
all_trips_v2 <- subset(all_trips, all_trips$ride_length > 0 
                       & all_trips$start_station_name != "HQ QR")
```

#### Here is the other possibility that I would like to compare for speed.
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR"
                            | all_trips$ride_length<0),]
                            
#### Just a quick check for oddities
```{r}
summary(all_trips_v2$ride_length)
```
This summary shows a large maximum value, so it is best to check for outliers.

```{r}
ride_length_outlier_check <- all_trips_v2 %>%
  arrange(desc(ride_length))
head(ride_length_outlier_check)
```
From this quick analysis, we see that there are multiple ride lengths over 2 million seconds. Note: in the above script, all the rider types were casual, which may be significant. MAKE A VISUAL TO ILLUSTRATE THIS!!! Show ride lengths, and frequency of certain bins---I think a histogram with separate columns for members and casual riders. Or, two separate histograms, kinda like a probability ditribution.

## Summary Statistics, Analysis, and Visualizations

#### Basic Overview: Mean, Median, Max, Min (In seconds)
```{r}
mean(all_trips_v2$ride_length)
median(all_trips_v2$ride_length)
max(all_trips_v2$ride_length)
min(all_trips_v2$ride_length)
sd(all_trips_v2$ride_length)
```

#### Summary Statistics, Separated by Rider Type
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = length)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = sd)
```

#### Number of Rides by Rider Type and Day of the Week (requested)
```{r}
## Number of rides by rider type and day of the week
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```
```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual))+
  geom_col(position = "dodge")

```

#### Ride Duration by Rider Type and Day of the Week
```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday, y=average_duration, fill=member_casual))+
  geom_col(position = "dodge")
```

#### Distribution of Ride Lengths (working version)
```{r}
all_trips_v2 %>%
  group_by(ride_length) %>%
  summarise(Rides=n()) %>%
ggplot(aes(x=ride_length, y = Rides)) +
  geom_col()
```

